OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2018-01-12 manuscript_nar.tex
More changes in R2

% Enter dates of publication
\copyrightyear{2017}
\pubdate{** ***** ****}
\pubyear{2017}
\jvolume{**}
\jissue{**}

@@

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 5~Kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla} (a single iteration of ICE), \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. When not specified, the
resolution of the analysis is 100~Kbp.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_1.eps}}
\caption{Modeling biases from the contact profile. A. Non linear
relationship between the number of restriction sites and the total number
of contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The purple line represents the raw signal, the dark blue
line represents the signal after OneD bias correction, the light blue line
represents the signal after OneD bias and copy number (CN) correction and
the orange line represents the signal after ICE bias correction. The long arm
of chromosome 17 (the region corresponding to 20-80 Mbp) is present in
four copies, explaining that the signal is about twice higher than for
the short arm. Both figures present data at 100~Kbp resolution}
\label{fig:totals}
\end{figure}

@@

In what follows, we benchmarked \textit{OneD} and \textit{OneD} with copy
number correction (\textit{OneD+CN}) against \textit{vanilla},
\textit{ICE} \citep{imakaev2012iterative}, the \textit{KR} algorithm
\citep{knight2013fast}, \textit{caICB} \citep{wu2016computational} and the
Local Genomic Features method \citep[\textit{LGF},][]{hu2012hicnorm,
servant2012hitc}. The first four methods correct biases implicitly,
whereas the fifth method does it explicitly.

@@

On the T47D data set, the methods that increased the reproducibility
between identical cell types also increased it between different cell
types (Figures~\ref{fig:aberrant}A), meaning that none of them enhanced
exclusively the biological signal. However, the difference was most marked
for \textit{OneD} and \textit{caICB}. On the K562 data set
(figure~\ref{fig:aberrant}B), both methods increased the reproducibility
between identical cell types but not between different cell types
%Figures~\ref{fig:aberrant}B)
, which is the desired behavior of a
correction method. On both data sets, \textit{OneD+CN}, \textit{ICE} and
\textit{KR} gave inferior results, while \textit{raw+CN} (copy number
correction alone), \textit{LGF} and \textit{vanilla} were not competitive.
The conclusions were unchanged when using other metrics to compare Hi-C
matrices (Supplementary Figures~4, 5 and 6).

@@

In this test, the reproducibility between different cell types was barely
affected (Figure~\ref{fig:normal}A). All the methods increased the
reproducibility between identical cell types (except one, see below), but
only when the experimental conditions were different. This means that all
the methods were able to remove some technical biases. \textit{ICE} and
\textit{KR} showed the strongest increase in reproducibility, but
\textit{OneD}, \textit{OneD+CN}, and \textit{caICB} had competitive
performance. \textit{vanilla} and \textit{LGF} were less competitive, and
\textit{raw+CN} (copy number correction alone) performed poorly, because
it is equivalent to no normalization on diploid cells. Other metrics gave
similar results (Supplementary Figures~8, 9 and 10, Supplementary Table~3).

@@

The results so far suggest that the copy number correction lowers the
performance of \textit{OneD}. A possible limitation in Hi-C is the absence
of ground truth. We thus used simulations to generate matrices with
defined karyotypic aberrations (see Materials and Methods) and tested the
capacity of the correction methods to distinguish different cell types
(Supplementary Figure~12). On this test, \textit{LGF} and \textit{OneD}
performed best. In comparison, \textit{OneD+CN} performed less well,
suggesting that correcting the copy number attenuates the biological
signal. This is corroborated by the fact that the copy number correction
alone has a lower performance than no correction at all. Thus, copy number
correction can indeed have side effects that lower the biological signal.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_4.eps}}
\caption{Copy number correction. A. Removing biases from Hi-C on breast
cancer cells. The reproducibility score between samples from the same
breast cancer cell type was computed after normalization. The vertical bar
shows the reproducibility score of the raw data. Shown are comparisons for
different (top) or identical (bottom) experimental conditions. B. Detail
of a Hi-C matrix normalized with different methods. The central portion
has an increased copy number, which affects normalization. \textit{ICE}
fades it away, \textit{LGF} enhances it and \textit{OneD} reduces the
signal by about half.}
\label{fig:cnv_correction}
\end{figure}

Copy number correction is also useful to give euploid-equivalent
representations of samples from aberrant karyotypes.
Figure~\ref{fig:cnv_correction}B shows an example where a region at the
center of the matrix is duplicated. \textit{ICE} overcompensated the
original bias and faded the signal almost entirely. Concomitantly, the
signal at the bottom left of the matrix was enhanced and showed a
structure that was not visible in the raw data. On the contrary,
\textit{LGF} strengthened the central region and the diagonal.
\textit{OneD+CN} reduced the level of the central portion by a factor 2
approximately, but did not otherwise distort the main features of the
region.

@@

\textit{OneD} corrects a single variable instead of the whole matrix, and
thus estimates the model parameters much faster than previous explicit
methods. We measured the running time of the different tools on a 3.3~GHz
machine with 62 GB~RAM, always using the default parameters.
Figure~\ref{fig:times}A shows the running times of the different methods
on all the samples shown in Supplementary Table~1 at 100~Kbp resolution.
The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
an over 100-fold span between the two.

>>>>

% Enter dates of publication
\copyrightyear{2018}
\pubdate{** ***** ****}
\pubyear{2018}
\jvolume{**}
\jissue{**}

@@

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 5~Kbp.
\textit{OneD} is implemented as an R package available at
http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla} (a single iteration of ICE), \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{HiCNorm} and
\textit{caICB} corrections respectively. When not specified, the
resolution of the analysis is 100~Kbp.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_1.eps}}
\caption{Modeling biases from the contact profile. A. Non linear
relationship between the number of restriction sites and the total number
of contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 9 of T47D. The purple line represents the raw signal, the dark blue
line represents the signal after OneD bias correction, the light blue line
represents the signal after OneD bias and copy number (CN) correction and
the orange line represents the signal after ICE bias correction. The grey
line represents the copy number estimation by independent CNV array and it
has been scaled to match the figure ranges. Some regions of the long arm of
chromosome 9 (the region corresponding to 60-140 Mbp) are present in four
copies, explaining that the signal is about twice higher than for the short
arm. Both figures present data at 100~Kbp resolution}
\label{fig:totals}
\end{figure}

@@

In what follows, we benchmarked \textit{OneD} and \textit{OneD} with copy
number correction (\textit{OneD+CN}) against \textit{vanilla},
\textit{ICE} \citep{imakaev2012iterative}, the \textit{KR} algorithm
\citep{knight2013fast}, \textit{caICB} \citep{wu2016computational} and the
Local Genomic Features method \citep[\textit{HiCNorm},][]{hu2012hicnorm,
servant2012hitc}. The first four methods correct biases implicitly,
whereas the fifth method does it explicitly.

@@

On the T47D data set, the methods that increased the reproducibility
between identical cell types also increased it between different cell
types (Figures~\ref{fig:aberrant}A), meaning that none of them enhanced
exclusively the biological signal. However, the difference was most marked
for \textit{OneD} and \textit{caICB}. On the K562 data set
(figure~\ref{fig:aberrant}B), both methods increased the reproducibility
between identical cell types but not between different cell types
%Figures~\ref{fig:aberrant}B)
, which is the desired behavior of a
correction method. On both data sets, \textit{OneD+CN}, \textit{ICE} and
\textit{KR} gave inferior results, while \textit{raw+CN} (copy number
correction alone), \textit{HiCNorm} and \textit{vanilla} were not competitive.
The conclusions were unchanged when using other metrics to compare Hi-C
matrices (Supplementary Figures~4, 5 and 6).

@@

In this test, the reproducibility between different cell types was barely
affected (Figure~\ref{fig:normal}A). All the methods increased the
reproducibility between identical cell types (except one, see below), but
only when the experimental conditions were different. This means that all
the methods were able to remove some technical biases. \textit{ICE} and
\textit{KR} showed the strongest increase in reproducibility, but
\textit{OneD}, \textit{OneD+CN}, and \textit{caICB} had competitive
performance. \textit{vanilla} and \textit{HiCNorm} were less competitive, and
\textit{raw+CN} (copy number correction alone) performed poorly, because
it is equivalent to no normalization on diploid cells. Other metrics gave
similar results (Supplementary Figures~8, 9 and 10, Supplementary Table~3).

@@

The results so far suggest that the copy number correction lowers the
performance of \textit{OneD}. A possible limitation in Hi-C is the absence
of ground truth. We thus used simulations to generate matrices with
defined karyotypic aberrations (see Materials and Methods) and tested the
capacity of the correction methods to distinguish different cell types
(Supplementary Figure~12). On this test, \textit{HiCNorm} and \textit{OneD}
performed best. In comparison, \textit{OneD+CN} performed less well,
suggesting that correcting the copy number attenuates the biological
signal. This is corroborated by the fact that the copy number correction
alone has a lower performance than no correction at all. Thus, copy number
correction can indeed have side effects that lower the biological signal.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_4.eps}}
\caption{Copy number correction. A. Removing biases from Hi-C on breast
cancer cells. The reproducibility score between samples from the same
breast cancer cell type was computed after normalization. The vertical bar
shows the reproducibility score of the raw data. Shown are comparisons for
different (top) or identical (bottom) experimental conditions. B. Detail
of a Hi-C matrix normalized with different methods. The central portion
has an increased copy number, which affects normalization. \textit{ICE}
fades it away, \textit{HiCNorm} enhances it and \textit{OneD} reduces the
signal by about half.}
\label{fig:cnv_correction}
\end{figure}

Copy number correction is also useful to give euploid-equivalent
representations of samples from aberrant karyotypes.
Figure~\ref{fig:cnv_correction}B shows an example where a region at the
center of the matrix is duplicated. \textit{ICE} overcompensated the
original bias and faded the signal almost entirely. Concomitantly, the
signal at the bottom left of the matrix was enhanced and showed a
structure that was not visible in the raw data. On the contrary,
\textit{HiCNorm} strengthened the central region and the diagonal.
\textit{OneD+CN} reduced the level of the central portion by a factor 2
approximately, but did not otherwise distort the main features of the
region.

@@

\textit{OneD} corrects a single variable instead of the whole matrix, and
thus estimates the model parameters much faster than previous explicit
methods. We measured the running time of the different tools on a 3.3~GHz
machine with 62 GB~RAM, always using the default parameters.
Figure~\ref{fig:times}A shows the running times of the different methods
on all the samples shown in Supplementary Table~1 at 100~Kbp resolution.
The fastest method was \textit{vanilla} and the slowest \textit{HiCNorm}, with
an over 100-fold span between the two.
